---
title: "Cyclistic Bike-Share Analysis"
author: "Soma Ganesh Dokku"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true          # adds a table of contents
    toc_float: true    # makes the TOC float on the side
    theme: cerulean    # Bootstrap theme (options: cerulean, cosmo, flatly, journal, lumen, paper, readable, sandstone, simplex, spacelab, united, yeti)
    highlight: tango   # syntax highlighting style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This report analyzes Cyclistic bike-share data using R.\
It mirrors the SQL workflow (import, preparation, cleaning, core analyses, advanced analyses, KPI dashboard) and adds visualizations for storytelling.

------------------------------------------------------------------------

# Setup

```{r load_libraries}
# Load libraries
library(tidyverse)
library(lubridate)
```

# List all CSV files in the working directory

```{r import_files}
files <- list.files(pattern = "*.csv")
```

# Read and combine them into one dataframe

```{r combine_data}
cyclistic_data <- files %>% map_df(~ read_csv(.x))
```

# Preview

```{r preview_data}
glimpse(cyclistic_data)
```

# Data Preparation

```{r data_preparation}
cyclistic_data <- cyclistic_data %>%
  mutate(
    started_at = lubridate::ymd_hms(started_at),
    ended_at   = lubridate::ymd_hms(ended_at),
    trip_duration_minutes = as.numeric(difftime(ended_at, started_at, units = "mins")),
    # Full month names, ordered Jan → Dec
    month = factor(lubridate::month(started_at, label = TRUE, abbr = FALSE),
                   levels = month.name, ordered = TRUE),
    # Day of week names
    day_of_week = lubridate::wday(started_at, label = TRUE, abbr = FALSE),
    # Seasons
    season = case_when(
      lubridate::month(started_at) %in% c(12,1,2) ~ "Winter",
      lubridate::month(started_at) %in% c(3,4,5) ~ "Spring",
      lubridate::month(started_at) %in% c(6,7,8) ~ "Summer",
      TRUE ~ "Fall"
    )
  )
```

# Data Cleaning

```{r data_cleaning}
cyclistic_data <- cyclistic_data %>%
  distinct(ride_id, .keep_all = TRUE) %>%
  filter(!is.na(started_at), !is.na(ended_at),
         !is.na(start_station_name), !is.na(end_station_name),
         trip_duration_minutes >= 0)
```

# Core Analyses

## Riders by Member Type

```{r riders_by_member_type}
cyclistic_data %>%
  count(member_casual) %>%
  ggplot(aes(x = member_casual, y = n, fill = member_casual)) +
  geom_col() +
  labs(title = "Riders by Member Type", y = "Number of Riders")
```

## Rides by Bike Type and Member Type

```{r rides_by_bike_type}
cyclistic_data %>%
  count(rideable_type, member_casual) %>%
  ggplot(aes(x = rideable_type, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Rides by Bike Type and Member Type", y = "Number of Riders")
```

## Monthly Riders

```{r monthly_riders}
cyclistic_data %>%
  count(month, member_casual) %>%
  ggplot(aes(x = month, y = n, color = member_casual, group = member_casual)) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "Monthly Riders by Member Type", y = "Number of Riders")
```

## Weekly Riders

```{r weekly_riders}
cyclistic_data %>%
  count(day_of_week, member_casual) %>%
  ggplot(aes(x = day_of_week, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Weekly Riders by Member Type", y = "Number of Riders")
```

## Hourly Riders

```{r hourly_riders}
cyclistic_data %>%
  mutate(hour = lubridate::hour(started_at)) %>%
  count(hour, member_casual) %>%
  ggplot(aes(x = hour, y = n, color = member_casual, group = member_casual)) +
  geom_line(size = 1.2) +
  labs(title = "Hourly Riders by Member Type", y = "Number of Riders")
```

# Advanced Analyses

## Ride Length Distribution

```{r ride_length_distribution}
ggplot(cyclistic_data, aes(x = trip_duration_minutes, fill = member_casual)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  xlim(0, 120) +
  labs(title = "Ride Length Distribution", x = "Duration (minutes)", y = "Count")
```

## Trip Duration

```{r trip_duration}
cyclistic_data %>%
  mutate(duration_bucket = case_when(
    trip_duration_minutes < 5 ~ "0–5 min",
    trip_duration_minutes <= 15 ~ "5–15 min",
    trip_duration_minutes <= 30 ~ "15–30 min",
    trip_duration_minutes <= 60 ~ "30–60 min",
    TRUE ~ "60+ min"
  )) %>%
  count(duration_bucket, member_casual) %>%
  ggplot(aes(x = duration_bucket, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Trip Duration Buckets", y = "Number of Rides")
```

## Seasonal Trends

```{r seasonal_trends}
cyclistic_data %>%
  count(season, member_casual) %>%
  ggplot(aes(x = season, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Seasonal Trends by Member Type", y = "Number of Rides")
```

## Station Analysis

```{r station_analysis}
cyclistic_data %>%
  count(start_station_name, sort = TRUE) %>%
  slice(1:10) %>%
  ggplot(aes(x = reorder(start_station_name, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Start Stations", y = "Number of Rides", x = "Station")
```

# KPI Dashboard

```{r kpi_dashboard}
total_rides <- nrow(cyclistic_data)

avg_duration <- mean(cyclistic_data$trip_duration_minutes)

peak_month <- cyclistic_data %>%
  count(month) %>%
  arrange(desc(n)) %>%
  slice(1)

member_split <- cyclistic_data %>%
  count(member_casual) %>%
  mutate(percentage = round(100 * n / sum(n), 2))

list(
  Total_Rides = total_rides,
  Avg_Duration_Minutes = avg_duration,
  Peak_Month = peak_month,
  Member_vs_Casual = member_split
)
```
